Future challenges for Research Software Engineering

### 2025-09-22 correspondence/rebuttal-1.md

### Suggestions

As suggested also by other referees, we made the workshop notes available at zenodo. A link has been made part of the paper.

> 1. The authors do not mention Germany as the focus of their discussion, probably because they want to address a broader aspect of RSEng that could apply to other countries or even globally. However, the history, focus and references clearly indicate a German perspective. I suggest either making it clear that the vision and discussion are limited to Germany, or discussing why the focus of the paper is rather nationalistic yet applies to a broader range of locations, and explaining where and why the stated aspects differ.

Thank you for highlighting this. We have added some text to the second paragraph of the introduction highlighting that with the paper focusing on the outputs of the German RSE conference, the thoughts and perspectives presented are largely German-focused. However, we also note that this work has broader relevance and we aim to highlight resources from the wider community where relevant.

### 2025-09-23 future_challenges_for_RSEng.tex
work out today's idea

% Article frontmatter
% ===================
\title{Future challenges for Research Software Engineering} % Title of the article
%\short{} % Short title of the article (optional)
\author{
Florian Goth\authorOrcid{0000-0003-2707-4790}\authorRef{1},
Leyla Jael Castro\authorOrcid{0000-0003-3986-0510}\authorRef{2},
Gerasimos Chourdakis\authorOrcid{0000-0002-3977-1385}\authorRef{3},
\texorpdfstring{\\}{} % line break
Simon Christ\authorOrcid{0000-0002-5866-1472}\authorRef{4},
Jeremy Cohen\authorOrcid{0000-0003-4312-2537}\authorRef{5},
Nicholas A. Del Grosso\thinspace\authorRef{6},
\texorpdfstring{\\}{} % line break
Jean-Noël Grad\authorOrcid{0000-0002-5821-4912}\authorRef{7},
Magnus Hagdorn\authorOrcid{0000-0002-5076-4864}\authorRef{8},
Toby Hodges\authorOrcid{0000-0003-1766-456X}\authorRef{9},
Jan Linxweiler\authorOrcid{0000-0002-2755-5087}\authorRef{10},
\texorpdfstring{\\}{} % line break
Frank Löffler\authorOrcid{0000-0001-6643-6323}\authorRef{11}\textsuperscript{,}\authorRef{12},
Michele Martone\authorOrcid{0000-0003-3239-8554}\authorRef{13},
Jan Philipp Thiele\authorOrcid{0000-0002-8901-6660}\authorRef{10}\textsuperscript{,}\authorRef{14}
} % Authors and references to addresses
\institute{% Institutes with labels
\autlabel{1} Institut für theoretische Physik 1, University of Würzburg, 97074, Würzburg, Germany\\
\autlabel{2} ZB MED Information Centre for Life Sciences, Cologne, Germany\\
\autlabel{3} Institute for Parallel and Distributed Systems, University of Stuttgart, Stuttgart, Germany\\
\autlabel{4} Leibniz University Hannover, Department of Cell Biology and Biophysics, Computational Biology, Germany\\
\autlabel{5} Imperial College London, London, UK\\
\autlabel{6} Institute for Experimental Epileptology and Cognition Research, Uniklinikum Bonn, Germany\\
\autlabel{7} Institute for Computational Physics, University of Stuttgart, Germany\\
\autlabel{8} Geschäftsbereich IT, Charité Universitätsmedizin Berlin, Germany\\
\autlabel{9} The Carpentries, USA\\
\autlabel{10} Technische Universität Braunschweig, Germany\\
\autlabel{11} Competence Center Digital Research, Friedrich Schiller University Jena, Germany\\
\autlabel{12} de-RSE e.V.---Society for Research Software in Germany\\
\autlabel{13} Independent researcher, Germany\\
\autlabel{14} Weierstrass Institute, Berlin, Germany;
              Leibniz University Hannover, Institute of Applied Mathematics, Scientific Computing, Hannover, Germany\\
}
\abstract{
It is 2025 and we are, admittedly, still working on establishing and growing Research Software Engineering (RSEng)
as a domain in its own right, with the hope of ensuring that it gets the recognition it deserves.
As the field develops, the role and value of Research Software Engineers (RSEs) becomes clearer.
With the growing importance and scope of software engineering in research,
researchers who code now begin to fill dedicated, specialized RSE roles crucial for the sustainable development
of long-term research software projects that can rapidly innovate.
Looking into the future, participants of the deRSE25 conference discussed how RSEng could evolve in response to the
rapid changes in the broader social, technological, and organisational landscape.
This paper presents the main discussion points of the conference participants,
imagining a future in which RSEs have a well-defined and well-integrated role in research institutions,
working as engineers of ever more complex research software projects,
taming a growing range of methods and tools,
and guiding researchers through technical and ethical considerations
related to research software.
} % Abstract of the article
\keywords{Future RSE, Evolving field, Vision} % Keywords for the article

@@

\subsection{How have academic Institutions transformed by then?}
\begin{whatis}{}{academic}
We expect research software to be a central part of scientific practice, on the same level as data and publications.
This will be supported by the first generations of RSE Master's graduates that find their way into research departments, and accelerate the pace of science.
Similar to RSEs, who are widely known and understood by university boards in 2035, we expect the concept of RSE units to be in the minds
of academic staff at any level.
We hope to see that universities actively use the presence of an RSE Department as a means to differentiate themselves from other universities.
We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
We also hope that specialised clusters form, which pool the skills of certain specialised RSEs with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
At universities which get such RSE units, we want to see that researchers naturally accept them as part of their available services.
Additionally, researchers will seamlessly interact with infrastructures for using research software.
\end{whatis}
Academia, while claiming to be innovative~\cite{wzvgbayern},
has organisationally often adopted time scales for change that are more reminiscent
to institutions thinking on the time scale of generations.
Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
hints at the potential for specialisation of research tasks.
This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
We have already seen some specialisation trends in recent years,
such as the HPC-RSE and AI-RSE, which are quite recent developments,
and whose appeal is stimulated by large public and private investments in HPC facilities and AI technologies.
We foresee new RSE specialisations emerging, \eg, in the fields of quantum computing,
such as the Quantum-RSE, or in digital research infrastructure.
We note the SE community has also independently expressed interest in creating an \emph{infrastructure engineering} role~\cite{Sochat2024Infra}
and funding bodies seem willing to promote such career pathways~\cite{UKRI2025Infra,EuroHPC2024Federated},
% European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
% European federation platform for HPC: https://eurohpc-ju.europa.eu/paving-way-eurohpc-federation-platform-2024-12-19_en
% European quantum computing: https://digital-strategy.ec.europa.eu/en/policies/quantum

@@

\begin{story}{Changes to the Universities}{changes}
Kim found a job years ago at one of the first RSE units in Arcadia. After a couple of years, Kim moved to the university of
Avalon which was interested in setting up their own department for RSEs in order to not fall behind the university of Arcadia.
Kim builds upon the model set forth by the University of Arcadia based on his valuable experience. Since the university
of Avalon has to follow different rules, not everything can be straightforwardly adapted. Life would be so much easier
if at least university regulations would be homogenised, no?
\end{story}

\subsection{Impact of Alternative Career Paths for Researchers on the Structure of Research Institutes}
\begin{whatis}{}{career-paths}
We would like to see the emergence of more dedicated staff, i.e. other (digital) RTPs next to the RSE.
Also we hope that researchers can build careers at the interface of academia, industry, and policy (\eg\ in think tanks, NGOs,
or organisations that do knowledge transfer).
\end{whatis}
RSEs can either be embedded in research groups (belonging to the mid-level university staff),
or be members of larger RSE units~\cite{Kempf2025-draft}.
The participants had the opinion that focusing on embedded RSEs would make onboarding new RSEs to existing projects easier.
This, however, comes with additional funding challenges,
as long-term funding can be more difficult to acquire in lower tiers
of a research organisation (i.e., at departmental or individual research group level).
Third-party funding can mitigate this issue in the mid-term.
However, the participants considered central RSE groups more likely in the future,
due to the fact that there are already examples and activities from various universities in this direction
(examples include the universities of Jena, Heidelberg, and Göttingen, as well as several universities
in the UK where the concept of dedicated RSE groups began to emerge around 2013).
We expect funding in central groups to evolve from initially temporary positions to permanent positions,
as the groups prove themselves to be increasingly valuable and, hence, sustainable.
Embedded RSEs and central groups will need to network with each other,
and potentially request and coordinate support from external RSEs.

\begin{story}{New career paths}{New career paths}
Ken, though trained as a physicist, found his passion not in theory,
but in keeping the quantum computers humming. He became the go-to expert
for troubleshooting entangled qubits and stabilizing delicate cryogenic systems.
His unique blend of physics knowledge and hands-on problem solving earned him
a new title among his peers: the first Quantum RTP.
In a world where theory meets reality, Ken quietly ensures that the future
of quantum computing stays on track.
\end{story}
\section{What is in this future for RSEs?}
%FG-NOTE: I think we can use this part to wrap things a bit up, and describe how this future then looks like.
%GC-NOTE: Shall we add another story box here? Ideas what we write here?
In the previous sections, we have tried to list some of the trends that we foresee.
The material presented here is based on the outputs of a workshop, and extended with the authors' ideas.
Nevertheless, there are some things that we declared out of scope here.
Among that is the ongoing digital transformation in wider society.
Digital competencies in the general population continue to increase,
making the jump to software development easier than before.
But this digitalisation will also lead to changes in the way societies
interact and therefore will translate into changes in how academia perceives itself.
We have also not sufficiently delved into the interaction with industry.
There are also some trends that we believe will stay with us over the coming years.
The differentiation of job profiles in academia, that has started with profiles such as the RSE will continue.
Societies will be facing new questions, arising from technologies such as generally available AI systems.

We hope that the standard of research software will have risen across the board,
enabling easier reuse of prior software outputs in new research.
We also hope that a cultural change is taking place to recognize research software
as an integral part of science that it is widely recognised and rewarded across
academia and is supported by political actors.
\section{Conclusion}
We have tried to elaborate on the state on the future ten years from now.
It turned out to be a difficult endeavour, since even the full breadth of today's trends is not yet foreseeable.
We have tried to shed some light on the social, technical and organisational aspects.
In summary, RSEs will very likely be doing different work to what they are doing today,
but their core task of enhancing the scientific output of researchers -- other human
beings -- will stay the same.
They will still develop software artefacts together with scientists,
since better software leads to better research,
and better research leads to better software.

>>>>

% Article frontmatter
% ===================
\title{Future challenges for Research Software Engineering} % Title of the article
%\short{} % Short title of the article (optional)
\author{
Florian Goth\authorOrcid{0000-0003-2707-4790}\authorRef{1},
Leyla Jael Castro\authorOrcid{0000-0003-3986-0510}\authorRef{2},
Gerasimos Chourdakis\authorOrcid{0000-0002-3977-1385}\authorRef{3},
\texorpdfstring{\\}{} % line break
Simon Christ\authorOrcid{0000-0002-5866-1472}\authorRef{4},
Jeremy Cohen\authorOrcid{0000-0003-4312-2537}\authorRef{5},
Nicholas A. Del Grosso\thinspace\authorRef{6},
\texorpdfstring{\\}{} % line break
Jean-Noël Grad\authorOrcid{0000-0002-5821-4912}\authorRef{7},
Magnus Hagdorn\authorOrcid{0000-0002-5076-4864}\authorRef{8},
Toby Hodges\authorOrcid{0000-0003-1766-456X}\authorRef{9},
Jan Linxweiler\authorOrcid{0000-0002-2755-5087}\authorRef{10},
\texorpdfstring{\\}{} % line break
Frank Löffler\authorOrcid{0000-0001-6643-6323}\authorRef{11},
Michele Martone\authorOrcid{0000-0003-3239-8554}\authorRef{12},
Jan Philipp Thiele\authorOrcid{0000-0002-8901-6660}\authorRef{10}\textsuperscript{,}\authorRef{13}
} % Authors and references to addresses
\institute{% Institutes with labels
\autlabel{1} Institut für theoretische Physik 1, University of Würzburg, 97074, Würzburg, Germany\\
\autlabel{2} ZB MED Information Centre for Life Sciences, Cologne, Germany\\
\autlabel{3} Institute for Parallel and Distributed Systems, University of Stuttgart, Stuttgart, Germany\\
\autlabel{4} Leibniz University Hannover, Department of Cell Biology and Biophysics, Computational Biology, Germany\\
\autlabel{5} Imperial College London, London, UK\\
\autlabel{6} Institute for Experimental Epileptology and Cognition Research, Uniklinikum Bonn, Germany\\
\autlabel{7} Institute for Computational Physics, University of Stuttgart, Germany\\
\autlabel{8} Geschäftsbereich IT, Charité Universitätsmedizin Berlin, Germany\\
\autlabel{9} The Carpentries, USA\\
\autlabel{10} Technische Universität Braunschweig, Germany\\
\autlabel{11} Michael Stifel Center Jena {\&} Friedrich Schiller University Jena, Germany\\
\autlabel{12} Independent researcher, Germany\\
\autlabel{13} Weierstrass Institute, Berlin, Germany;
              Leibniz University Hannover, Institute of Applied Mathematics, Scientific Computing, Hannover, Germany\\
}
\abstract{
It is 2025 and the research software community is, admittedly, still working on establishing and growing Research Software Engineering (RSEng)
as a domain in its own right, with the hope of ensuring that it gets the recognition it deserves.
As the field develops, the role and value of Research Software Engineers (RSEs) becomes clearer.
With the growing importance and scope of software engineering in research,
researchers who code now begin to fill dedicated, specialized RSE roles crucial for the sustainable development
of long-term research software projects that can rapidly innovate.
Looking into the future, participants of the deRSE25 conference discussed how RSEng could evolve in response to the
rapid changes in the broader social, technological, and organisational landscape.
This paper presents the main discussion points of the conference participants,
imagining a future in which RSEs have a well-defined and well-integrated role in research institutions,
working as engineers of ever more complex research software projects,
taming a growing range of methods and tools,
and guiding researchers through technical and ethical considerations
related to research software.
} % Abstract of the article
\keywords{Future RSE, Evolving field, Vision} % Keywords for the article

@@

\subsection{Transformations in Academic Institutions}
\begin{whatis}{}{academic}
Research software is a central part of scientific practice, on the same level as data and publications.
This will be supported by the first generations of RSE Master's graduates that find their way into research departments, and accelerate the pace of science.
Similar to RSEs, who are widely known and understood by university boards in 2035, academic staff members at any level are aware of the concept of RSE units.
Universities actively use the presence of an RSE Department as a means to differentiate themselves from other universities.
Some central RSE units will have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
Specialised clusters have formed, which pool the skills of certain specialised RSEs with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
At universities which create such RSE units, it can be observed that researchers naturally accept them as part of their available services.
Additionally, researchers will seamlessly interact with infrastructures for using research software.
\end{whatis}
Academia, while claiming to be innovative~\cite{wzvgbayern},
has organisationally often adopted time scales for change that are more reminiscent
to institutions thinking on the time scale of generations.
Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
hints at the potential for specialisation of research tasks.
This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
Some specialisation trends have already been observed in recent years,
such as the HPC-RSE and AI-RSE, which are quite recent developments,
and whose appeal is stimulated by large public and private investments in HPC facilities and AI technologies.
We foresee new RSE specialisations emerging, \eg, in the fields of quantum computing,
such as the Quantum-RSE, or in digital research infrastructure.
We note the SE community has also independently expressed interest in creating an \emph{infrastructure engineering} role~\cite{Sochat2024Infra}
and funding bodies seem willing to promote such career pathways~\cite{UKRI2025Infra,EuroHPC2024Federated},
% European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
% European federation platform for HPC: https://eurohpc-ju.europa.eu/paving-way-eurohpc-federation-platform-2024-12-19_en
% European quantum computing: https://digital-strategy.ec.europa.eu/en/policies/quantum

@@

\begin{story}{Changes to the Universities}{changes}
Kim found a job years ago at one of the first RSE units in Arcadia. After a couple of years, Kim moved to the university of
Avalon which was interested in setting up their own department for RSEs in order to not fall behind the university of Arcadia.
Kim builds upon the model set forth by the University of Arcadia based on his valuable experience. Since the university
of Avalon has to follow different legal and administrative rules, not everything can be straightforwardly adapted.
Life would be so much easier if at least university regulations were homogenised.
\end{story}

\subsection{Impact of Alternative Career Paths for Researchers on the Structure of Research Institutes}
\begin{whatis}{}{career-paths}
As the RSE becomes established, we are also seeing the emergence of a wider range of more specialized, technical roles within the research community, i.e. other (digital) RTPs.
The legal requirements for employing people have become flexible enough to allow for the employment of more RTPs.
Work outside of the publication record is more readily accepted and enables researchers to build careers at the interface of academia, industry, and policy
(\eg\ in think tanks, NGOs, or organisations that do knowledge transfer).
\end{whatis}
RSEs can either be embedded in research groups (belonging to the mid-level university staff),
or be members of larger RSE units~\cite{Kempf2025-draft}.
The participants had the opinion that focusing on embedded RSEs would make onboarding new RSEs to existing projects easier.
This, however, comes with additional funding challenges,
as long-term funding can be more difficult to acquire in lower tiers of a research organisation (i.e., at departmental or individual research group level).
Third-party funding can mitigate this issue in the mid-term.
However, the participants considered central RSE groups more likely in the future,
due to the fact that there are already examples and activities from various universities in this direction
(examples include the universities of Jena, Heidelberg, and Göttingen, as well as several universities
in the UK where the concept of dedicated RSE groups began to emerge around 2013).
We expect funding in central groups to evolve from initially temporary positions to permanent positions,
as the groups prove themselves to be increasingly valuable and, hence, sustainable.
Embedded RSEs and central groups will need to network with each other,
and potentially request and coordinate support from external RSEs.
Within Germany at least, the distinction around whether these positions are classed as research staff or technical staff has so far created some discussion.
The authors hope that, 10 years from now, this discussion has resolved towards a solution that enables the employment of RTPs more flexibly,
and thereby recognizing that the distinction between research staff and technical staff is too restrictive.
A similar discussion is currently ongoing in the UK with initiatives like \emph{the Technician's Commitment}~\cite{techniciancommitment}, and \emph{the Hidden Ref}~\cite{hiddenref},
and finding a resolution with having new job titles, pay grades, and career options according to the level of work that you do.
The authors' expect these initiatives to spread and find a wider resonance over the world till 2035.

\begin{story}{New career paths}{New career paths}
Ken, though trained as a physicist, found his passion not in theory,
but in keeping the quantum computers humming. He became the go-to expert
for troubleshooting entangled qubits and stabilizing delicate cryogenic systems.
His unique blend of physics knowledge and hands-on problem solving earned him
a new title among his peers: the first Quantum RTP.
In a world where theory meets reality, Ken quietly ensures that the future
of quantum computing stays on track.
\end{story}

\section{What is in this Future for RSEs?}
%FG-NOTE: I think we can use this part to wrap things a bit up, and describe how this future then looks like.
%GC-NOTE: Shall we add another story box here? Ideas what we write here?
In the previous sections, we have tried to list some of the trends that we foresee.
The material presented here is based on the outputs of a workshop, and extended with the authors' ideas.
Nevertheless, there are some things that we declared out of scope, such as the ongoing digital transformation in wider society.
Digital competencies in the general population continue to increase,
making the jump to software development easier than before.
But this digitalisation will also lead to changes in the way societies
interact and therefore will translate into changes in how academia perceives itself.
We have also not sufficiently delved into the interaction with industry.
There are also some trends that we believe will stay with us over the coming years.
The differentiation of job profiles in academia, that has started with profiles such as the RSE will continue.
Societies will be facing new questions, arising from technologies such as generally available AI systems.

We hope that the standard of research software will have risen across the board,
enabling easier reuse of prior software outputs in new research.
We also hope that a cultural change is taking place to recognize research software
as an integral part of science that it is widely recognised and rewarded across
academia and is supported by political actors.

\section{Conclusion}
We have tried to elaborate on the state of the future ten years from now.
It turned out to be a difficult endeavour, since even the full breadth of today's trends is not yet foreseeable.
We have tried to shed some light on the social, technical and organisational aspects.
In summary, RSEs will very likely be doing different work to what they are doing today,
but their core task of enhancing the scientific output of researchers -- other human
beings -- will stay the same.
They will still develop software artefacts together with scientists,
since better software leads to better research,
and better research leads to better software.
